---
title: 'Phase 4: Logistic Regression Model'
author: "Masheia Dzimba and Peter Mangoro"
date: "2025-12-02"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(pROC)
library(caret)
library(knitr)
```

# Introduction

This document presents Phase 4: Logistic Regression Model. We build a logistic regression model to predict 30-day hospital readmissions, interpret coefficients and odds ratios, and evaluate model performance.

# Load Data

```{r load-data, echo=FALSE, comment=""}
load("data_logistic.RData")
cat("Dataset: ", nrow(data_logistic), " observations, ", ncol(data_logistic), " variables\n")
```
# Train/Test Split

```{r train-test, echo=FALSE, comment=""}
set.seed(1)
train_indices <- createDataPartition(data_logistic$readmitted, 
                                     p = 0.7, 
                                     list = FALSE)

data_train <- data_logistic[train_indices, ]
data_test <- data_logistic[-train_indices, ]

cat("Training set: ", nrow(data_train), " observations (70%)\n")
cat("Testing set: ", nrow(data_test), " observations (30%)\n")
cat("Readmission rate - Training: ", round(mean(data_train$readmitted)*100, 2), "%\n")
cat("Readmission rate - Testing: ", round(mean(data_test$readmitted)*100, 2), "%\n")
```

# Build Model

```{r build-model,comment=""}
# Build model formula
predictor_vars <- setdiff(colnames(data_train), "readmitted")
formula_str <- paste("readmitted ~", paste(predictor_vars, collapse = " + "))
formula_obj <- as.formula(formula_str)

# Fit the model
model_logistic <- glm(formula_obj, 
                     data = data_train, 
                     family = binomial(link = "logit"))

```

# Logistic Regression Equation

## Mathematical Formulation

The logistic regression model uses the logistic function to model the probability of readmission:


$$P(\text{Readmitted} = 1 \mid X) = \frac{e^{\beta_0 + \sum_{i=1}^{p} \beta_i X_i}}{1 + e^{\beta_0 + \sum_{i=1}^{p} \beta_i X_i}}$$


Or equivalently, 

$$\text{logit}(P) = \ln\left(\frac{P}{1-P}\right) = \beta_0 + \sum_{i=1}^{p} \beta_i X_i$$

Where:

- $P(\text{Readmitted} = 1 \mid X)$ = Probability of readmission given predictor variables

- $\beta_0$ = Intercept (baseline log-odds)

- $\beta_1, \beta_2, ..., \beta_p$ = Coefficients for each predictor variable

- $X_1, X_2, ..., X_p$ = Predictor variables

## Fitted Model Equation

```{r fitted-equation, echo=FALSE, comment="", results='asis'}
# Extract intercept and top coefficients
intercept <- round(coef(model_logistic)[1], 4)
all_coefs <- coef(model_logistic)[-1]  # All except intercept
top_coefs <- head(sort(abs(all_coefs), decreasing = TRUE), 5)
top_vars <- names(top_coefs)
top_coef_values <- round(all_coefs[top_vars], 4)


# Build equation with line breaks using split environment
# Break after every 2 terms for better viewing
cat("$$\\begin{split}\n")
cat("\\text{logit}(P) &= ", intercept, sep = "")

# Process terms in groups of 2
terms_per_line <- 2
line_num <- 1
for(i in 1:length(top_vars)) {
  var_name <- gsub("_", "\\\\_", top_vars[i])
  
  # Break line after every 2 terms (except for the first line)
  if(i > 1 && (i - 1) %% terms_per_line == 0) {
    cat(" \\\\\n&\\quad")
  }
  
  if(top_coef_values[i] >= 0) {
    cat(" + ", top_coef_values[i], " \\cdot X_{\\text{", var_name, "}}", sep = "")
  } else {
    cat(" ", top_coef_values[i], " \\cdot X_{\\text{", var_name, "}}", sep = "")
  }
}

cat(" + \\text{... (other predictors)}\n")
cat("\\end{split}$$\n\n")

cat("**Where:**\n\n")
cat("* $P$ = Probability of readmission\n")
cat("* $\\beta_0 = ", intercept, "$ (intercept)\n", sep = "")

for(i in 1:length(top_vars)) {
  var_name <- gsub("_", "\\\\_", top_vars[i])
  cat("* $\\beta_{\\text{", var_name, "}} = ", top_coef_values[i], "$\n", sep = "")
}
```



## Example Calculation

```{r example-calculation, echo=FALSE, results='asis'}
# Create example patient values (using top 3 predictors)
example_vars <- top_vars[1:min(3, length(top_vars))]
example_coefs <- top_coef_values[1:min(3, length(top_coef_values))]

# Get example values from actual dataset or use custom values
example_values <- numeric(length(example_vars))
names(example_values) <- example_vars

for(i in 1:length(example_vars)) {
  var <- example_vars[i]
  
  # Check if custom value is provided
  if(exists("custom_values") && !is.null(custom_values) && var %in% names(custom_values)) {
    example_values[i] <- custom_values[var]
  } else if(var %in% colnames(data_train)) {
    # For numerical variables, use median
    if(is.numeric(data_train[[var]])) {
      example_values[i] <- round(median(data_train[[var]], na.rm = TRUE), 2)
    } else {
      # For categorical/dummy variables (0/1), use 1 to show the effect
      example_values[i] <- 1
    }
  } else {
    # If variable not found, use a reasonable default
    example_values[i] <- 3
  }
}


# TO CUSTOMIZE VALUES: Comment or Uncomment and modify the lines below
# example_values["n_diagnoses"] <- 3
 example_values["n_inpatient"] <- 2
# example_values["medspec_medical_specialtyEmergency.Trauma"] <- 1

cat("For a patient with:\n\n")
for(i in 1:length(example_vars)) {
  var_name <- gsub("_", " ", example_vars[i])
  cat("* ", var_name, " = ", example_values[i], "\n", sep = "")
}

# Calculate logit using actual values
example_logit <- intercept + sum(example_coefs * example_values)
example_prob <- round(exp(example_logit) / (1 + exp(example_logit)), 4)

cat("\n**Logit calculation:**\n\n")

# Build logit equation with line breaks using actual values
cat("$$\\begin{split}\n")
cat("\\text{logit}(P) &= ", intercept, sep = "")
for(i in 1:length(example_vars)) {
  coef <- example_coefs[i]
  val <- example_values[i]
  if(coef >= 0) {
    cat(" + ", coef, " \\cdot ", val, sep = "")
  } else {
    cat(" ", coef, " \\cdot ", val, sep = "")
  }
}
cat(" \\\\\n&= ", round(example_logit, 4), "\n", sep = "")
cat("\\end{split}$$\n\n")

cat("**Probability calculation:**\n\n")
prob_string <- paste0("P = \\frac{e^{", round(example_logit, 4), "}}{1 + e^{", round(example_logit, 4), "}} = ", example_prob)
cat("$$", prob_string, "$$\n\n", sep = "")

cat("**Interpretation:** This patient has a ", round(example_prob * 100, 2), "% probability of readmission.\n", sep = "")
```


## Coefficient Interpretation

**Coefficients ($\beta$):**

- **Sign**: Positive coefficients increase the log-odds (and probability) of readmission; negative coefficients decrease it

- **Magnitude**: Larger absolute values indicate stronger effects


**Example**: If $\beta_{\text{n\_inpatient}} = 0.38$:

- A one-unit increase in previous inpatient visits increases the log-odds of readmission by 0.38

- This corresponds to an odds ratio of $e^{0.38} = 1.46$ (46% increase in odds)

## Gradient (Rate of Change)

The **gradient** represents how quickly the probability changes with respect to each predictor:

$$\frac{\partial P}{\partial X_i} = \beta_i \cdot P(1-P)$$

**Key Points:**

- The gradient is **not constant** - it depends on the current probability $P$

- Maximum gradient occurs when $P = 0.5$ (steepest part of the S-curve)

- The gradient is smaller when $P$ is close to 0 or 1 (flatter parts of the curve)

# Model Summary

```{r model-summary, echo=FALSE, comment=""}
model_summary <- summary(model_logistic)
print(model_summary)
```

# Regression Output

```{r regression-output, echo=FALSE, comment=""}
# Extract coefficients
coefficients_table <- model_summary$coefficients

# Create comprehensive output table
output_table <- data.frame(
  Variable = rownames(coefficients_table),
  Coefficient = round(coefficients_table[, "Estimate"], 4),
  Std_Error = round(coefficients_table[, "Std. Error"], 4),
  Z_Value = round(coefficients_table[, "z value"], 4),
  P_Value = round(coefficients_table[, "Pr(>|z|)"], 6),
  Odds_Ratio = round(exp(coefficients_table[, "Estimate"]), 4),
  CI_Lower = round(exp(coefficients_table[, "Estimate"] - 1.96 * coefficients_table[, "Std. Error"]), 4),
  CI_Upper = round(exp(coefficients_table[, "Estimate"] + 1.96 * coefficients_table[, "Std. Error"]), 4)
)

# Split into multiple tables for better PDF display
significant_vars <- output_table[output_table$P_Value < 0.05, ]
significant_vars_ordered <- significant_vars[order(significant_vars$P_Value), ]

# Table 1: Top 5 Most Significant (with full details)
top5_table <- head(significant_vars_ordered, 5)
kable(top5_table[, c("Variable", "Coefficient", "Std_Error", "Z_Value", "P_Value")],
      caption = "Top 5 Most Significant Variables: Coefficients and Statistics",
      digits = 4, row.names = FALSE)

# Table 2: Top 5 Odds Ratios
top5_odds <- head(significant_vars_ordered, 5)
kable(top5_odds[, c("Variable", "Odds_Ratio", "CI_Lower", "CI_Upper", "P_Value")],
      caption = "Top 5 Most Significant Variables: Odds Ratios and Confidence Intervals",
      digits = 4, row.names = FALSE)

# Table 3: Next 10 Significant Variables (summary)
if(nrow(significant_vars_ordered) > 5) {
  next10_table <- significant_vars_ordered[6:min(15, nrow(significant_vars_ordered)), ]
  kable(next10_table[, c("Variable", "Coefficient", "Odds_Ratio", "P_Value")],
        caption = "Additional Significant Variables (p < 0.05)",
        digits = 4, row.names = FALSE)
}
```

# Hypothesis Testing

```{r hypothesis-testing, echo=FALSE, comment="", results='asis'}
cat("**Hypothesis Testing**\n\n")

cat("For each variable in the model:\n\n")
cat("$H_0$: $\\beta_{\\text{variable}} = 0$ (variable has no effect on readmission)\n\n")
cat("$H_1$: $\\beta_{\\text{variable}} \\neq 0$ (variable affects readmission)\n\n")

# Show top significant variables
top_significant <- head(significant_vars[order(significant_vars$P_Value), ], 5)
kable(top_significant[, c("Variable", "Coefficient", "P_Value", "Odds_Ratio")],
      caption = "Top 5 Most Significant Variables", row.names = FALSE)
```

# R-squared Interpretation

```{r rsquared,echo=FALSE,comment="", results='asis'}

# Calculate pseudo R-squared
null_deviance <- model_summary$null.deviance
residual_deviance <- model_summary$deviance
mcfadden_r2 <- 1 - (residual_deviance / null_deviance)

cat("McFadden's Pseudo $R^2$: ", round(mcfadden_r2, 4), "\n\n")
cat("Interpretation: The model explains approximately ", 
    round(mcfadden_r2 * 100, 2), 
    "% of the variance in readmission status.\n")
```

# Model Evaluation

```{r model-evaluation,echo=FALSE,comment=""}

# Make predictions
predictions_prob <- predict(model_logistic, newdata = data_test, type = "response")
predictions_class <- ifelse(predictions_prob > 0.5, 1, 0)

# Confusion Matrix
conf_matrix <- table(Predicted = predictions_class, Actual = data_test$readmitted)
kable(conf_matrix, caption = "Confusion Matrix")

# Calculate metrics
TN <- conf_matrix[1, 1]
FP <- conf_matrix[2, 1]
FN <- conf_matrix[1, 2]
TP <- conf_matrix[2, 2]

accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
specificity <- TN / (TN + FP)
f1_score <- 2 * (precision * recall) / (precision + recall)

metrics_table <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall (Sensitivity)", "Specificity", "F1-Score"),
  Value = c(accuracy, precision, recall, specificity, f1_score),
  Percentage = c(accuracy * 100, precision * 100, recall * 100, specificity * 100, f1_score * 100)
)

kable(metrics_table, caption = "Model Performance Metrics", digits = 2)
```

# ROC Curve

```{r roc-curve, fig.cap="ROC Curve: Logistic Regression Model", echo=FALSE,comment=""}
roc_obj <- roc(data_test$readmitted, predictions_prob)
auc_value <- auc(roc_obj)

cat("Area Under the Curve (AUC): ", round(as.numeric(auc_value), 4), "\n")

# Create ROC plot
roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

p_roc <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", linewidth = 1) +
  labs(title = "ROC Curve: Logistic Regression Model",
       subtitle = paste("AUC =", round(as.numeric(auc_value), 4)),
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))
print(p_roc)
```

# Odds Ratios Interpretation

```{r odds-ratios, echo=FALSE, comment=""}

cat("Top Variables with Highest Odds Ratios (Risk Factors):\n\n")

# Split into two tables
top_odds <- output_table[order(-output_table$Odds_Ratio), ]
top_odds_filtered <- top_odds[top_odds$P_Value < 0.05, ]  # Only significant ones

# Table 1: Top 5 Risk Factors
top5_risk <- head(top_odds_filtered, 5)
kable(top5_risk[, c("Variable", "Odds_Ratio", "CI_Lower", "CI_Upper", "P_Value")],
      caption = "Top 5 Risk Factors (Highest Odds Ratios)",
      digits = 4, row.names = FALSE)

# Table 2: Next 5 Risk Factors
if(nrow(top_odds_filtered) > 5) {
  next5_risk <- top_odds_filtered[6:min(10, nrow(top_odds_filtered)), ]
  kable(next5_risk[, c("Variable", "Odds_Ratio", "CI_Lower", "CI_Upper", "P_Value")],
        caption = "Additional Risk Factors",
        digits = 4, row.names = FALSE)
}
```

# Summary

```{r summary, echo=FALSE, comment="", results='asis'}
# Calculate summary statistics dynamically
# Accuracy (from model evaluation)
predictions_prob_summary <- predict(model_logistic, newdata = data_test, type = "response")
predictions_class_summary <- ifelse(predictions_prob_summary > 0.5, 1, 0)
conf_matrix_summary <- table(Predicted = predictions_class_summary, Actual = data_test$readmitted)
TN_summary <- conf_matrix_summary[1, 1]
FP_summary <- conf_matrix_summary[2, 1]
FN_summary <- conf_matrix_summary[1, 2]
TP_summary <- conf_matrix_summary[2, 2]
accuracy_summary <- (TP_summary + TN_summary) / (TP_summary + TN_summary + FP_summary + FN_summary)

# AUC (from ROC)
roc_obj_summary <- roc(data_test$readmitted, predictions_prob_summary)
auc_value_summary <- as.numeric(auc(roc_obj_summary))

# Pseudo R² (from rsquared section)
null_deviance_summary <- model_summary$null.deviance
residual_deviance_summary <- model_summary$deviance
mcfadden_r2_summary <- 1 - (residual_deviance_summary / null_deviance_summary)

# Number of significant variables
n_significant <- sum(output_table$P_Value < 0.05)

# Top predictor (highest odds ratio among significant variables)
top_odds_summary <- output_table[order(-output_table$Odds_Ratio), ]
top_odds_filtered_summary <- top_odds_summary[top_odds_summary$P_Value < 0.05, ]
if(nrow(top_odds_filtered_summary) > 0) {
  top_predictor_name <- top_odds_filtered_summary$Variable[1]
  top_predictor_or <- round(top_odds_filtered_summary$Odds_Ratio[1], 2)
  # Clean up variable name for display
  top_predictor_display <- gsub("_", " ", top_predictor_name)
  top_predictor_display <- gsub("medspec medical specialty", "medical specialty", top_predictor_display)
} else {
  top_predictor_display <- "N/A"
  top_predictor_or <- 0
}

cat("- **Accuracy**: ", round(accuracy_summary * 100, 2), "%\n", sep = "")
cat("- **AUC**: ", round(auc_value_summary, 3), "\n", sep = "")
cat("- **Pseudo $R^2$**: ", round(mcfadden_r2_summary * 100, 2), "%\n", sep = "")
cat("- **", n_significant, " significant variables** identified (p < 0.05)\n", sep = "")
cat("- **Top predictor**: ", top_predictor_display, " (OR: ", top_predictor_or, ")\n", sep = "")
```


