---
title: "Phase 6: Model Comparison - Logistic Regression vs. CART"
author: "Masheia Dzimba and Peter Mangoro"
date: "2025-12-06"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Install packages if needed
if (!require(tidyr)) {
  install.packages("tidyr")
  library(tidyr)
}

library(ggplot2)
library(dplyr)
library(pROC)
library(knitr)
```

# Introduction

This document presents Phase 6: Model Comparison. We compare the performance of Logistic Regression, CART, and Random Forest models, analyze their interpretability, and justify model selection.

# Load Results

```{r load-results, echo=FALSE, comment=""}
# Load performance metrics from all models
metrics_logistic <- read.csv("plots/04_performance_metrics.csv")
metrics_cart <- read.csv("plots/05_performance_metrics.csv")
metrics_rf <- read.csv("plots/05b_performance_metrics.csv")

auc_logistic <- read.csv("plots/04_auc.csv")
auc_cart <- read.csv("plots/05_auc.csv")
auc_rf <- read.csv("plots/05b_auc.csv")

regression_output <- read.csv("plots/04_regression_output.csv")
var_importance_cart <- read.csv("plots/05_variable_importance.csv")
var_importance_rf <- read.csv("plots/05b_variable_importance.csv")

# Load Random Forest model metadata (if available)
if(file.exists("plots/05b_model_metadata.csv")) {
  model_metadata_rf <- read.csv("plots/05b_model_metadata.csv")
  n_trees_rf <- model_metadata_rf$Value[model_metadata_rf$Parameter == "ntree"]
} else {
  n_trees_rf <- 500  # Default value
}
```

# Performance Metrics Comparison

```{r performance-comparison, echo=FALSE, comment=""}
# Create comparison table with all three models
comparison_table <- data.frame(
  Metric = metrics_logistic$Metric,
  Logistic_Regression = round(metrics_logistic$Percentage, 2),
  CART = round(metrics_cart$Percentage, 2),
  Random_Forest = round(metrics_rf$Percentage, 2)
)

# Add AUC row
comparison_table <- rbind(comparison_table,
  data.frame(
    Metric = "AUC",
    Logistic_Regression = round(auc_logistic$Value * 100, 2),
    CART = round(auc_cart$Value * 100, 2),
    Random_Forest = round(auc_rf$Value * 100, 2)
  )
)

kable(comparison_table, caption = "Side-by-Side Performance Comparison", row.names = FALSE)
```

# Visualization

## Performance Metrics Comparison

```{r metrics-viz, fig.cap="Model Performance Comparison", echo=FALSE, comment=""}
comparison_data <- data.frame(
  Model = c("Logistic Regression", "CART", "Random Forest"),
  AUC = c(auc_logistic$Value, auc_cart$Value, auc_rf$Value),
  Accuracy = c(metrics_logistic$Percentage[metrics_logistic$Metric == "Accuracy"] / 100,
               metrics_cart$Percentage[metrics_cart$Metric == "Accuracy"] / 100,
               metrics_rf$Percentage[metrics_rf$Metric == "Accuracy"] / 100),
  Precision = c(metrics_logistic$Percentage[metrics_logistic$Metric == "Precision"] / 100,
                metrics_cart$Percentage[metrics_cart$Metric == "Precision"] / 100,
                metrics_rf$Percentage[metrics_rf$Metric == "Precision"] / 100),
  Recall = c(metrics_logistic$Percentage[metrics_logistic$Metric == "Recall (Sensitivity)"] / 100,
             metrics_cart$Percentage[metrics_cart$Metric == "Recall (Sensitivity)"] / 100,
             metrics_rf$Percentage[metrics_rf$Metric == "Recall (Sensitivity)"] / 100),
  F1_Score = c(metrics_logistic$Value[metrics_logistic$Metric == "F1-Score"],
               metrics_cart$Value[metrics_cart$Metric == "F1-Score"],
               metrics_rf$Value[metrics_rf$Metric == "F1-Score"])
)

comparison_long <- comparison_data %>%
  select(Model, Accuracy, Precision, Recall, F1_Score) %>%
  pivot_longer(cols = c(Accuracy, Precision, Recall, F1_Score),
               names_to = "Metric",
               values_to = "Value")

p_metrics <- ggplot(comparison_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Logistic Regression" = "steelblue", 
                               "CART" = "lightcoral",
                               "Random Forest" = "darkgreen")) +
  labs(title = "Model Performance Comparison",
       subtitle = "Accuracy, Precision, Recall, and F1-Score",
       y = "Score",
       x = "Metric",
       fill = "Model") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12))
print(p_metrics)
```

## AUC Comparison

```{r auc-comparison, fig.cap="AUC Comparison: Logistic Regression vs. CART", echo=FALSE,comment=""}
p_auc <- ggplot(comparison_data, aes(x = Model, y = AUC, fill = Model)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("Logistic Regression" = "steelblue", 
                               "CART" = "lightcoral",
                               "Random Forest" = "darkgreen")) +
  labs(title = "AUC Comparison: All Three Models",
       y = "Area Under the Curve (AUC)",
       x = "Model") +
  geom_text(aes(label = round(AUC, 4)), vjust = -0.5, size = 5) +
  ylim(0, max(comparison_data$AUC) * 1.2) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        legend.position = "none")
print(p_auc)
```

# Interpretability Comparison

```{r interpretability, echo=FALSE, comment="", results='asis'}
cat("**Logistic Regression:**\n\n")
cat("- Provides coefficients and odds ratios for each variable\n\n")
cat("- Statistical significance testing (p-values)\n\n")
cat("- ", nrow(regression_output), " parameters in the model\n\n", sep = "")
cat("- ", sum(regression_output$P_Value < 0.05, na.rm = TRUE), " significant variables (p < 0.05)\n\n", sep = "")

cat("**CART:**\n\n")
cat("- Simple decision tree with ", nrow(var_importance_cart), " variables considered\n\n", sep = "")
cat("- Very interpretable: simple decision rule(s)\n\n")
cat("- Non-linear relationships captured\n\n")
cat("- Top variable: ", var_importance_cart$Variable[1], " (", 
    round(var_importance_cart$Importance_Percent[1], 2), "% importance)\n\n", sep = "")

cat("**Random Forest:**\n\n")
cat("- Ensemble of 500 decision trees\n\n")
cat("- Uses bootstrap sampling and feature randomization\n\n")
cat("- ", nrow(var_importance_rf), " variables considered\n\n", sep = "")
cat("- Lower interpretability (ensemble effect)\n\n")
cat("- Non-linear relationships captured\n\n")
cat("- Top variable: ", var_importance_rf$Variable[1], " (", 
    round(var_importance_rf$Importance_Percent[1], 2), "% importance)\n", sep = "")
```

# Model Selection

```{r model-selection, echo=FALSE, comment="", results='asis'}

# Determine winner for each metric
lr_wins <- sum(comparison_table$Logistic_Regression > comparison_table$CART & 
               comparison_table$Logistic_Regression > comparison_table$Random_Forest, na.rm = TRUE)
cart_wins <- sum(comparison_table$CART > comparison_table$Logistic_Regression & 
                 comparison_table$CART > comparison_table$Random_Forest, na.rm = TRUE)
rf_wins <- sum(comparison_table$Random_Forest > comparison_table$Logistic_Regression & 
               comparison_table$Random_Forest > comparison_table$CART, na.rm = TRUE)

cat("Performance Metrics Won:\n")
cat("  Logistic Regression: ", lr_wins, " metrics\n")
cat("  CART: ", cart_wins, " metrics\n")
cat("  Random Forest: ", rf_wins, " metrics\n\n")

# Overall assessment - find best AUC and accuracy
auc_values <- c(auc_logistic$Value, auc_cart$Value, auc_rf$Value)
names(auc_values) <- c("Logistic Regression", "CART", "Random Forest")
best_auc_model <- names(auc_values)[which.max(auc_values)]

acc_values <- c(metrics_logistic$Value[metrics_logistic$Metric == "Accuracy"],
                metrics_cart$Value[metrics_cart$Metric == "Accuracy"],
                metrics_rf$Value[metrics_rf$Metric == "Accuracy"])
names(acc_values) <- c("Logistic Regression", "CART", "Random Forest")
best_acc_model <- names(acc_values)[which.max(acc_values)]

if(best_auc_model == best_acc_model) {
  best_model <- best_auc_model
  reason <- paste("Highest AUC and accuracy")
} else if(max(auc_values) > 0.7 && max(acc_values) > 0.6) {
  best_model <- best_auc_model
  reason <- paste("Highest AUC (", round(max(auc_values), 3), ")", sep = "")
} else {
  best_model <- "Logistic Regression"
  reason <- "Best overall performance and statistical rigor"
}

cat("**Recommended Model: ", best_model, "**\n")
cat("**Reason**: ", reason, "\n\n")

# Create justification table directly (instead of reading from CSV)
justification <- data.frame(
  Criterion = c("Accuracy", "AUC", "Precision", "Recall", "F1-Score", 
                "Interpretability", "Complexity", "Statistical Rigor"),
  Logistic_Regression = c(
    paste0(round(metrics_logistic$Percentage[metrics_logistic$Metric == "Accuracy"], 2), "%"),
    paste0(round(auc_logistic$Value * 100, 2), "%"),
    paste0(round(metrics_logistic$Percentage[metrics_logistic$Metric == "Precision"], 2), "%"),
    paste0(round(metrics_logistic$Percentage[metrics_logistic$Metric == "Recall (Sensitivity)"], 2), "%"),
    round(metrics_logistic$Value[metrics_logistic$Metric == "F1-Score"], 4),
    "High (coefficients, odds ratios)",
    paste0("High (", nrow(regression_output), " parameters)"),
    "High (p-values, hypothesis tests)"
  ),
  CART = c(
    paste0(round(metrics_cart$Percentage[metrics_cart$Metric == "Accuracy"], 2), "%"),
    paste0(round(auc_cart$Value * 100, 2), "%"),
    paste0(round(metrics_cart$Percentage[metrics_cart$Metric == "Precision"], 2), "%"),
    paste0(round(metrics_cart$Percentage[metrics_cart$Metric == "Recall (Sensitivity)"], 2), "%"),
    round(metrics_cart$Value[metrics_cart$Metric == "F1-Score"], 4),
    "Very High (simple tree, easy rules)",
    "Very Low (simple tree)",
    "Medium (no p-values, variable importance)"
  ),
  Random_Forest = c(
    paste0(round(metrics_rf$Percentage[metrics_rf$Metric == "Accuracy"], 2), "%"),
    paste0(round(auc_rf$Value * 100, 2), "%"),
    paste0(round(metrics_rf$Percentage[metrics_rf$Metric == "Precision"], 2), "%"),
    paste0(round(metrics_rf$Percentage[metrics_rf$Metric == "Recall (Sensitivity)"], 2), "%"),
    round(metrics_rf$Value[metrics_rf$Metric == "F1-Score"], 4),
    paste0("Low (ensemble of ", n_trees_rf, " trees)"),
    paste0("High (", n_trees_rf, " trees, complex ensemble)"),
    "Medium (variable importance, no p-values)"
  )
)

kable(justification, caption = "Model Comparison: Detailed Justification", row.names = FALSE)
```

# Key Findings

```{r key-findings, echo=FALSE, comment="", results='asis'}

# Calculate average accuracy
avg_accuracy <- mean(c(metrics_logistic$Percentage[metrics_logistic$Metric == "Accuracy"],
                       metrics_cart$Percentage[metrics_cart$Metric == "Accuracy"],
                       metrics_rf$Percentage[metrics_rf$Metric == "Accuracy"]))

cat("1. **Performance:**\n")
cat("   - All three models show similar performance (accuracy ~", round(avg_accuracy, 0), "%)\n", sep = "")
cat("   - Best AUC: ", best_auc_model, " (", round(max(auc_logistic$Value, auc_cart$Value, auc_rf$Value), 3), ")\n", sep = "")
cat("   - Logistic Regression: ", round(auc_logistic$Value, 3), 
    ", CART: ", round(auc_cart$Value, 3), 
    ", Random Forest: ", round(auc_rf$Value, 3), "\n", sep = "")
max_auc <- max(auc_logistic$Value, auc_cart$Value, auc_rf$Value)
cat("   - All models have ", 
    ifelse(max_auc < 0.7, "fair to poor", "good"), 
    " discrimination (AUC < 0.7)\n\n")

cat("2. **Interpretability:**\n")
n_cart_vars <- nrow(var_importance_cart)
n_lr_params <- nrow(regression_output)
n_rf_vars <- nrow(var_importance_rf)
cat("   - CART: Simplest (", n_cart_vars, " variables, single tree)\n", sep = "")
cat("   - Logistic Regression: ", n_lr_params, " parameters, detailed statistical insights\n", sep = "")
cat("   - Random Forest: Most complex (", n_trees_rf, " trees, ", n_rf_vars, " variables)\n", sep = "")
cat("   - All models identify similar key predictors\n\n")

cat("3. **Key Predictors:**\n")
# Get top predictor from logistic regression
top_lr_predictor <- regression_output[order(-regression_output$Odds_Ratio), ][1, ]
cat("   - Logistic Regression: ", top_lr_predictor$Variable, " (OR: ", round(top_lr_predictor$Odds_Ratio, 2), "), age groups, medical specialty\n", sep = "")
cat("   - CART: ", var_importance_cart$Variable[1], " (", 
    round(var_importance_cart$Importance_Percent[1], 2), "% importance)\n", sep = "")
cat("   - Random Forest: ", var_importance_rf$Variable[1], " (", 
    round(var_importance_rf$Importance_Percent[1], 2), "% importance)\n", sep = "")
```

# Summary

```{r summary, echo=FALSE, comment="", results='asis'}
# Calculate summary statistics dynamically
lr_wins_summary <- sum(comparison_table$Logistic_Regression > comparison_table$CART & 
                       comparison_table$Logistic_Regression > comparison_table$Random_Forest, na.rm = TRUE)
cart_wins_summary <- sum(comparison_table$CART > comparison_table$Logistic_Regression & 
                         comparison_table$CART > comparison_table$Random_Forest, na.rm = TRUE)
rf_wins_summary <- sum(comparison_table$Random_Forest > comparison_table$Logistic_Regression & 
                       comparison_table$Random_Forest > comparison_table$CART, na.rm = TRUE)
total_metrics <- nrow(comparison_table)

cat("This phase compared all three models:\n\n")
cat("- **Logistic Regression** wins ", lr_wins_summary, " out of ", total_metrics, " performance metrics\n", sep = "")
cat("- **CART** offers superior simplicity and interpretability\n")
cat("- **Random Forest** wins ", rf_wins_summary, " out of ", total_metrics, " performance metrics\n", sep = "")
cat("- **Recommended**: ", best_model, " (", reason, ")\n", sep = "")
cat("- All models identify **", var_importance_cart$Variable[1], "** as a key predictor\n", sep = "")
```

